Abstract Binding Trees

### 2015-10-29 paper.tex
Rearrange semantic universe; start defining interpretation

\newcommand\Cats{\mathbf{Cat}}
\newcommand\Discrete[1]{#1_{\equiv}}
\newcommand\Sets{\mathbf{Set}}
\newcommand\SCtx{\mathbf{SCtx}}
\newcommand\Ctx{\mathbf{Ctx}}
\newcommand\Seq{\mathbf{Seq}}

\newcommand\OpCat[1]{#1^{op}}

@@

\newcommand\Define[2]{\IMode{#1}\triangleq\OMode{#2}}

\DeclareMathOperator\Yoneda{\mathbf{y}}

@@

Let $\FinSets$ be the free cocartesian category on a single object (that is,
the category of finite cardinals and functions between them); then the comma
construction $\Define{\Ctx}{\Comma{\FinSets}{\Sorts}}$ is the category of
variable contexts, and $\Define{\Seq}{\SCtx\times\Ctx\times\Sorts}$ is the category of
sequents $\Upsilon\parallel\Gamma\vdash s$.  Then, we fix the presheaf category $\Sets^\Seq$ as
our semantic universe.

\subsection{Signatures and substitution monoidal structures}

Any two presheaves $A,B : \Sets^\Seq$ have a tensor $A\bullet B$,
defined as a coend in the following way:
\[
  \Define{(A\bullet B)(\Upsilon\parallel\Gamma\vdash s)}{
    \int^{\Delta\in\Ctx}
      A(\Upsilon\parallel\Delta\vdash s)
        \otimes \prod_{x\in\Dom\Delta} B(\Upsilon\parallel\Gamma\vdash\Delta(x))
  }
\]

The presheaf of variables $\Define{V(\Upsilon\parallel\Gamma\vdash s)}{\MkSet{x \in \Dom\Gamma
\mid \Gamma(x) = s}}$ is the unit to this tensor, and together they comprise a
substitution monoidal structure. Then, a presheaf $A$ is a monoid if it is
equipped with the following natural transformations:

\[
  \begin{tikzcd}[ampersand replacement = \&]
    {\IMode{V}} \arrow[r, "\IMode\nu"]
    \& {\IMode{A}}
    \& {\IMode{A\bullet A}}
    \arrow[l, "\IMode\varsigma", swap]
  \end{tikzcd}
\]

Intuitively, $\nu$ embeds variables into $A$ and $\varsigma$ equips $A$ with an
operation for simultaneous substitutions of variables.

>>>>

\newcommand\Cats{\mathbf{Cat}}
\newcommand\Discrete[1]{#1_{\equiv}}
\newcommand\Sets{\mathbf{Set}}
\newcommand\SCtx{\mathbf{SCtx}}
\newcommand\Ctx{\mathbf{Ctx}}
\newcommand\Seq{\mathbf{Seq}}
\newcommand\HypCat{\mathbf{H}}

\newcommand\OpCat[1]{#1^{\mathsf{op}}}

@@

\newcommand\Define[2]{\IMode{#1}\triangleq\OMode{#2}}
\newcommand\ADefine[2]{\IMode{#1}&\triangleq\OMode{#2}}

\newcommand\SigFun[1]{\mathcal{F}_{#1}}

\newcommand\SemBrackets[1]{\left\llbracket #1\right\rrbracket}
\newcommand\Yoneda[1]{\boldsymbol{y}(#1)}

@@

Let $\FinSets$ be the free cocartesian category on a single object (that is,
the category of finite cardinals and functions between them); then the comma
construction $\Define{\Ctx}{\Comma{\FinSets}{\Sorts}}$ is the category of
variable contexts. Let $\Define{\HypCat}{\OpCat{(\SCtx\times\Ctx)}}$; then we
fix the functor category $\widehat{\HypCat}^\Sorts$ as our semantic universe.

\subsection{Substitution monoidal structures}

For a presheaf $A:\widehat{\HypCat}$ and a sort-indexed family of presheaves
$P:\widehat{\HypCat}^\Sorts$, we have an operation $A\bullet P$,
defined as a coend in the following way:
\[
  \Define{(A\bullet P)(\Upsilon\parallel\Gamma)}{
    \int^{\Delta\in\OpCat\Ctx}
      A(\Upsilon\parallel\Delta)
      \times \prod_{x\in\Dom\Delta} P_{\Delta(x)}(\Upsilon\parallel\Gamma)
  }
\]

Using this, we can define a tensor $P\odot Q$ for
$\Of{P,Q}{\widehat{\HypCat}^\Sorts}$ as follows:
\[
  \HypJ{
    \Define{(P\odot Q)_s}{
      P_s\bullet Q
    }
    \quad
  }{
    \Member{s}{\Sorts}
  }
\]

Then, variables object
$\Define{V_s(\Upsilon\parallel\Gamma)}{\MkSet{x\in\Dom\Gamma\mid \Gamma(x)=s}}$
is the unit to this tensor. We will say that an object
$\Of{P}{\widehat{\HypCat}^\Sorts}$ is a $\Sigma$-monoid in case it is equipped
with the following natural transformations:
\[
  \begin{tikzcd}[ampersand replacement = \&]
    {\IMode{V}} \arrow[r, "\IMode\nu"]
    \& {\IMode{P}}
    \& {\IMode{P\odot P}}
    \arrow[l, "\IMode\varsigma", swap]
  \end{tikzcd}
\]
Intuitively, $\nu$ embeds variables into $P$ and $\varsigma$ equips $P$ with an
operation for simultaneous substitutions of variables.

\begin{remark}
  A \emph{relative monad} on a functor $\Of{J}{\mathbf{J}\to\mathbf{C}}$ is a
  monoid in the functor category $\mathbf{C}^\mathbf{J}$
  \cite{altenkirch-chapman-uustalu:2015}. A presheaf $P$ with a substitution
  monoidal structure $\Pair{\nu}{\varsigma}$ can be viewed as a relative monad
  on the functor $\Of{V}{\widehat{\HypCat}^\Sorts}$ of variables.
\end{remark}

For any context $\Gamma$, we have an induced map
$\Of{\nu_\Gamma}{\left(\prod_{x\in\Dom\Gamma}V_{\Gamma(x)}\right)\to
\left(\prod_{x\in\Dom\Gamma}P_{\Gamma(x)}\right)}$.

\subsection{The signature endofunctor and its initial algebras}

For each signature $\Match{\Sigma}{\Pair{\Sorts}{\Operators}}$, we have an
endofunctor
$\Of{\SigFun{\Sigma}}{\widehat{\HypCat}^\Sorts\to\widehat{\HypCat}^\Sorts}$,
which is defined as follows:

\begin{align*}
  \IMode{\SigFun{\Sigma}(X)_s(\Upsilon\parallel\Gamma)}
    &\triangleq
    \OMode{
      \coprod_{\vartheta\in\Operators\Pair{\Upsilon}{\MkArity{\vec{v}}{s}}}
      \prod_{\MkValence{\vec{p}}{\vec{q}}{s'}\in\vec{v}}
      X_{s'}(\Upsilon,\vec{p}\parallel\Gamma,\vec{q})
    }
\end{align*}

\begin{remark}
  By $\Upsilon,\vec{p}$ we mean any extension of $\Upsilon$ by fresh symbols $\vec{u}$ of sorts $\vec{p}$.
\end{remark}

Then, a $\Sigma$-model is a $\Sigma$-monoid $P$ which is equipped with an initial
algebra $\Of{\alpha}{\SigFun{\Sigma}(P)\to P}$, which shall interpret
applications of each operator.

\subsection{Interpretations of terms}

\[
  \Define{
    \SemBrackets{
      \Omega\triangleright\Upsilon\parallel\Gamma
    }_P
  }{
    \left(\prod_{(\MV{m}:\MkValence{\vec{p}}{\vec{q}}{s})\in\Omega}
    P_{s}^{\Yoneda{\vec{p}\parallel\vec{q}})}\right)
    \times
    \prod_{x\in\Dom\Gamma}V_{\Gamma(x)}
  } %% ???
\]

Let $\Of{\SemBrackets{\Omega\triangleright\Upsilon\parallel\Gamma\vdash
M:s}_P}{\SemBrackets{\Omega\triangleright\Upsilon\parallel\Gamma}_P\to P_s}$ be
defined as follows:
\begin{align*}
  \ADefine{
    \SemBrackets{\Omega\triangleright\Upsilon\parallel\Gamma\vdash x_j : s}_P
  }{
    \begin{tikzcd}[ampersand replacement = \&]
      {\SemBrackets{\Omega\triangleright\Upsilon\parallel\Gamma}_P} \arrow[r, "\pi_2"]
      \& { \prod_{x\in\Dom\Gamma}V_{\Gamma(x)} } \arrow[r, "\nu_{\Gamma}"]
      \& {\prod_{x\in\Dom\Gamma} P_{\Gamma(x)}} \arrow[r, "\pi_j"]
      \& P_s
    \end{tikzcd}
  }
\end{align*}
